The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum

### 2014-04-01 cordillera-climate.tex
Add corrections from Irina in manuscript

where $\phi$ is given in degrees and $z$ in meters above current sea level. This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.

Sea level is lowered by 120\,m and basal topography responds to ice load following a bedrock deformation model that includes point-wise isostasy, elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite half-space \citep{lingle-clark-1985,bueler-etal-2007}. It uses a lithosphere density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times 10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times 10^{21}$\,\unit{Pa\,s}. Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response. Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation.

@@

Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature. This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse rate of -6\,\unit{{\degree}C\,km^{-1}} is used in all simulations. No corrections apply to precipitation changes with elevation.

@@

Because most of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalysis datasets, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of -6\,\unit{{\degree}C\,km^{-1}} as in the simulations. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, which may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}). However significant disparity among reanalyses exists, too. More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).

@@

Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin more closely (Fig.~\ref{fig:best}) do not reach a steady-state. Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load. In turn, a~key assumption for the present study is our choice of 10\,kyr for the duration of all our simulations. Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,kyr, they provide no lower bound to this value.

@@

Our representation of climate history by constant temperature offsets is voluntarily simplistic. To better understand the transient character of the LGM Cordilleran ice sheet, time-dependent palaeo-climate forcing would be needed. Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, their patterns are not trivial and may potentially display a~strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets. As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model. Although the ice sheet interior likely became dryer, precipitation may increased along part of the margin where the presence of ice imposed ascending winds. In this context, using a~GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may help producing more westerly-centred ice configurations than those presented here (Fig.~\ref{fig:best}).

Furthermore, the surface mass balance model crudely simplifies the processes taking place at the glacier surface. Refreezing of melted snow and ice and water retention in the snow pack was shown to influence surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model. Based on scarce observations over the Greenland ice sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}. However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and it remains undemonstrated whether their application to a~different region and time period is sound. In addition, we simulate temperature variability by using a~constant, uniform value of daily temperature standard deviation. This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.

The bedrock deformation model uses homogeneous lithospheric properties and a semi-infinite mantle of constant viscosity. Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include influence from the neighbouring Laurentide ice sheet. These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.

@@

% fig:tempheatmap
\begin{figure}
  \includegraphics{cordillera-climate-tempheatmap}
  \caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis. Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of -6\,\unit{{\degree}C\,km^{-1}}. Colour mapping is based on a~logarithmic scale. Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
  \label{fig:tempheatmap}
\end{figure}

% fig:tempdiff
\begin{figure}
  \includegraphics{cordillera-climate-tempdiff}
  \caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation and lapse-rate correction. Note the cold bias of NCEP/NCAR data and temperature anomalies due to unresolved topographic detail.}
  \label{fig:tempdiff}
\end{figure}

% fig:precheatmap
\begin{figure}
  \includegraphics{cordillera-climate-precheatmap}
  \caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis, after bi-linear spatial interpolation. Colour mapping is based on a~logarithmic scale. Note the wet bias of all reanalysis data relative to WorldClim data.}
  \label{fig:precheatmap}
\end{figure}

% fig:precdiff
\begin{figure}
  \includegraphics{cordillera-climate-precdiff}
  \caption{Winter (DJF) precipitation rate difference maps against WorldClim data, after bi-linear spatial interpolation. Note the wet bias of all reanalysis data and the large anomalies of CFSR and NCEP/NCAR data.}
  \label{fig:precdiff}
\end{figure}

>>>>

where $\phi$ is given in degrees and $z$ in meters above current sea level. This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.

Sea level is lowered by 120\,m and basal topography responds to ice load following a bedrock deformation model that includes point-wise isostasy, elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite half-space \citep{lingle-clark-1985,bueler-etal-2007}. It uses a lithosphere density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times 10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times 10^{21}$\,\unit{Pa\,s}. Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response.

@@

Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature. This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all simulations. No corrections apply to precipitation changes with elevation.

@@

Because much of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalyses, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, such as positive anomalies in the northernmost part of the modelling domain. These may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}). However, significant disparity among reanalyses exists, too. More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).

@@

Most of the simulations presented in this study, and more particularly those that reproduce the geomorphological LGM margin more closely (Fig.~\ref{fig:best}) do not reach a steady-state. Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load. In turn, a~key assumption for the present study is our choice of 10\,kyr for the duration of all our simulations. Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,kyr, they provide no lower bound to this value.

@@

Our representation of climate history by constant temperature offsets is voluntarily simplistic. To better understand the transient character of the LGM Cordilleran ice sheet, time-dependent palaeo-climate forcing would be needed. Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, their patterns are not trivial and may potentially display a~strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets. As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model. These changes likely consist of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds. They could result in a more westerly-centred ice sheet than modelled here, with lower ice thickness in its interior. In addition, the final position of the eastern ice margin is largely controlled by its advance rate through the run. Therefore, the precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model. Although using a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may represent a first step towards including ice sheet feedback on climate, their spatial resolution do not allow for accurate modelling of orographic precipitation changes in a region as mountainous as the North American Cordillera.

Furthermore, the surface mass balance model crudely simplifies the processes taking place at the glacier surface. Refreezing of melted snow and ice and water retention in the snow pack was shown to influence surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model. Based on scarce observations over the Greenland ice sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}. However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and it remains undemonstrated whether their application to a~different region and time period is sound. In addition, we simulate temperature variability by using a~constant, uniform value of daily temperature standard deviation. This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rogozhina-rau-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.

The bedrock deformation model uses homogeneous lithospheric properties and a semi-infinite mantle of constant viscosity. Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include the influence from the neighbouring Laurentide ice sheet. These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.

@@

% fig:tempheatmap
\begin{figure}
  \includegraphics{cordillera-climate-tempheatmap}
  \caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis. Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of 6\,\unit{{\degree}C\,km^{-1}}. Colour mapping is based on a~logarithmic scale. Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
  \label{fig:tempheatmap}
\end{figure}

% fig:tempdiff
\begin{figure}
  \includegraphics{cordillera-climate-tempdiff}
  \caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bilinear spatial interpolation and lapse-rate correction. Note the cold bias of NCEP/NCAR data and temperature anomalies due to unresolved topographic detail.}
  \label{fig:tempdiff}
\end{figure}

% fig:precheatmap
\begin{figure}
  \includegraphics{cordillera-climate-precheatmap}
  \caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis, after bilinear spatial interpolation. Colour mapping is based on a~logarithmic scale. Note the wet bias of all reanalysis data relative to WorldClim data.}
  \label{fig:precheatmap}
\end{figure}

% fig:precdiff
\begin{figure}
  \includegraphics{cordillera-climate-precdiff}
  \caption{Winter (DJF) precipitation rate difference maps against WorldClim data, after bilinear spatial interpolation. Note the wet bias of all reanalysis data and the large anomalies of CFSR and NCEP/NCAR data.}
  \label{fig:precdiff}
\end{figure}
